---
title: "Statistical Analysis of Monthly Background Checks of Gun Purchases"
author: "Gregor Aisch, Josh Keller and Dirk Eddelbuettel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Monthly Background Checks of Gun Purchases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
library(gunsales)
```

## About

This document provides the analysis from the [New York Times](http://www.nytimes.com) 
on-line article ["What Drives Gun Sales: Terrorism, Obama and Calls for Restrictions"](http://www.nytimes.com/interactive/2015/12/10/us/gun-sales-terrorism-obama-restrictions.html?). The [R](http://www.r-project.org) code underlying the analysis is provided
in the [GitHub](http://www.github.com) repository [gunsales](https://github.com/NYTimes/gunsales) which contains the R package 


## Part I Data

The first step consists in transforming the two raw data sets included in the
package into the data.frame used for the subsequent plotting.

```{r, initialDataFake, eval=FALSE}
gunsales <- analysis()
```

```{r, initialData, fig.width=7, echo=FALSE}
if (gunsales:::.goodOS()) gunsales <- analysis() else cat("Unsupported platform -- no plots below.")
```

## Part II: Base Plots

Given the transformed data, we can display the variety of plots contained in
the New York Times analysis (which uses post-processed variants suitable for
publication).

```{r, basePlotsFake, eval=FALSE}
plot_gunsales(gunsales)
```

```{r, basePlots, fig.width=7, echo=FALSE}
if (gunsales:::.goodOS()) plot_gunsales(gunsales)
```

## Part III: Using ggplot

The second set of charts redisplays the same charts as before, but using the
[ggplot2](http://cloud.r-project.org/package=ggplot2) package.

```{r, ggPlotsFake, eval=FALSE}
ggplot_gunsales(gunsales)
```

```{r, ggPlots, fig.width=7, echo=FALSE}
if (gunsales:::.goodOS()) ggplot_gunsales(gunsales)
```
